Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-02-17 alloscore_manuscript/alloscore-application-shorter.tex
convert supplement to appendices

Of the 10 models evaluated for this one week, the CU-select model had the most accurate forecasts according to the
allocation score while the USC-SI\_kJalpha model had the most accurate forecasts based on MWIS (Table
\ref{tab:multi-k-scores}). The JHUAPL-Bucky model had the second best MWIS but the third worst allocation score.

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Feb 16 20:39:10 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  per-capita & 865 & - & 1029 & 366 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{For one illustrative week, a comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS), (see Section \ref{sec:ias_examp}). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

A state-wise comparison using Figures \ref{fig:thermometer-plot} and \ref{fig:multi-loc-ranks}A of the JHUAPL-Bucky and
CU-select models shows that while the JHUAPL-Bucky forecast distributions were more centered on the eventual
observations in many states, the suggested allocations of the JHUAPL-Bucky model were less efficient than those of the
CU-select model. Thus JHUAPL-Bucky achieved a worse allocation score than CU-select (on this particular week) by
allocating excess resources to several states, such as Pennsylvania and Michigan, rather than to states such as Florida
and California where these resources would have reduced unmet need.  These allocation errors of JHUAPL-Bucky resulted
from forecasts failing to consistently capture the relative resource needs across different states. The CU-select model
made some similar errors — most prominently, over-allocating resources to Ohio — but overall, it more successfully
forecasted the resource demands across different locations in relative terms.

On the other hand, CU-select had worse performance as measured by MWIS. Its forecasts were biased downwards, and it
consistently incurred a large penalty for underprediction (Figure \ref{fig:multi-loc-ranks}B). The JHUAPL-Bucky model
had wider predictive intervals which more often included the observed level of daily hospital admissions. Its MWIS was
therefore less severely penalized by under- or over-predictions of the actual hospitalization counts.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/multi-loc-ranks-1} \caption[Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select)]{Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the `observed resource need' bar being larger than the `allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.}\label{fig:multi-loc-ranks}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/metrics-over-time-1} \caption[Hospital admissions and evaluation metrics over time]{Hospital admissions and evaluation metrics over time. Panel A shows the number of hospital admissions in the US as a whole due to COVID-19 on a sequence of 15 Mondays from December 2021 through March 2022. These are the dates for which forecasts were made and evaluated. A horizontal dashed line at 15,000 shows the hypothetical resource constraint $K$. Panel B shows allocation scores (AS) for each model's 14 day-ahead forecast, across all US states. The $x$-axis corresponds to the date of the observation that a model's prediction was targeting (e.g., the date the forecast was made plus the forecast horizon). AS typically are high when the observed value is near to the constraint, which occurs during the last Monday in December (on the way up) and the last Monday in January (on the way down). In Panel B, the per-capita allocation is drawn in a heavier solid line. Panel C shows the MWIS metric across weeks, averaged across all states. MWIS values tend to scale with the observed and predicted values, and the peak MWIS values happen around and just after the peak of the Omicron wave.}\label{fig:metrics-over-time}
\end{figure}

@@

Averaging across all weeks, the ensemble forecast achieved a better mean allocation score (MAS) than two benchmark
methods. The COVIDhub-ensemble had the best MAS across all weeks, the per-capita allocation had the second best MAS, and
the COVIDhub-baseline model had the sixth best allocation (Table \ref{tab:multi-week-performance-summary}). No
individual model had a better MAS than the per-capita allocation.

Model rankings on mean allocation score (MAS) and time-averaged mean weighted interval score
(taMWIS) were broadly similar, with some places of disagreement. The four most accurate and the five least accurate models were the same
according to both metrics, although not in exactly the same order (Table \ref{tab:multi-week-performance-summary}).
(This comparison excludes the per-capita allocation which does not use forecasts and therefore cannot be assigned a MWIS.)
Notably, the JHUAPL-SLPHospEns model ranked best on taMWIS but had only the fourth best MAS.

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Feb 16 20:39:15 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
model & MAS & taMWIS & taMWIS rank \\ 
  \hline
COVIDhub-ensemble & 389 & 70 & 2 \\ 
  per-capita & 464 & - & - \\ 
  COVIDhub-trained\_ensemble & 483 & 87 & 4 \\ 
  CU-select & 502 & 81 & 3 \\ 
  JHUAPL-SLPHospEns & 526 & 67 & 1 \\ 
  COVIDhub-baseline & 594 & 93 & 5 \\ 
  JHUAPL-Bucky & 643 & 112 & 7 \\ 
  MUNI-ARIMA & 707 & 116 & 8 \\ 
  JHUAPL-Gecko & 929 & 110 & 6 \\ 
  USC-SI\_kJalpha & 1473 & 155 & 9 \\ 
   \hline
\end{tabular}
\caption{Time averaged mean weighted interval scores (taMWIS) and mean allocation scores (MAS) by model across 13 weeks. These results show the average performance across time for the nine models that submitted forecasts for every week from 2021-11-29 through 2022-02-21, as well
  as for a per-capita allocation. MWIS is not defined for the per-capita allocation which does not use forecasts.} 
\label{tab:multi-week-performance-summary}
\end{table}

@@

AS was computed for a range of $K$ from 200 to 60,000 for forecasts made on December 20, 2020 predicting levels of
hospitalizations on January 3, 2021 as well as for the per-capita allocation (Figure \ref{fig:multi-k}A). These
calculations highlight that AS was highest at at values of $K$ near the observed nationwide total hospital admissions of
19,581 that day. Model ranks by AS were fairly stable across all $K$, and there appears
to be an substantial interval around the observed value over which ranks are constant.

@@

While our example used a hypothetical scenario to illustrate a potential application of the allocation score, resource
allocation decisions are a realistic example of the kinds of decisions that could motivate more targeted forecasting
exercises.
%One real-world example, which was the motivating example presented in the work above, is allocation of ventilators during a respiratory viral pandemic\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}.

>>>>

Of the 10 models evaluated for this one week, the CU-select model had the most accurate forecasts according to the
allocation score while the USC-SI\_kJalpha model had the most accurate forecasts based on MWIS (Table
\ref{tab:multi-k-scores}). The \texttt{JHUAPL-Bucky} model had the second best MWIS but the third worst allocation score.

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Sat Feb 17 02:22:59 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  per-capita & 865 & - & 1029 & 366 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{For one illustrative week, a comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS), (see Section \ref{sec:ias_examp}). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

A state-wise comparison using Figures \ref{fig:thermometer-plot} and \ref{fig:multi-loc-ranks}A of the \texttt{JHUAPL-Bucky} and
\texttt{CU-select} models shows that while the \texttt{JHUAPL-Bucky} forecast distributions were more centered on the eventual
observations in many states, the suggested allocations of the JHUAPL-Bucky model were less efficient than those of the
\texttt{CU-select} model. Thus \texttt{JHUAPL-Bucky} achieved a worse allocation score than CU-select (on this particular week) by
allocating excess resources to several states, such as Pennsylvania and Michigan, rather than to states such as Florida
and California where these resources would have reduced unmet need.  These allocation errors of \texttt{JHUAPL-Bucky} resulted
from forecasts failing to consistently capture the relative resource needs across different states. The \texttt{CU-select} model
made some similar errors — most prominently, over-allocating resources to Ohio — but overall, it more successfully
forecasted the resource demands across different locations in relative terms.

On the other hand, \texttt{CU-select} had worse performance as measured by MWIS. Its forecasts were biased downwards, and it
consistently incurred a large penalty for underprediction (Figure \ref{fig:multi-loc-ranks}B). The \texttt{JHUAPL-Bucky} model
had wider predictive intervals which more often included the observed level of daily hospital admissions. Its MWIS was
therefore less severely penalized by under- or over-predictions of the actual hospitalization counts.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/multi-loc-ranks-1} \caption[Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (\texttt{JHUAPL-Bucky} and \texttt{CU-select})]{Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (\texttt{JHUAPL-Bucky} and \texttt{CU-select}). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the `observed resource need' bar being larger than the `allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.}\label{fig:multi-loc-ranks}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/metrics-over-time-1} \caption[Hospital admissions and evaluation metrics over time]{Hospital admissions and evaluation metrics over time. Panel A shows the number of hospital admissions in the US as a whole due to COVID-19 on a sequence of 15 Mondays from December 2021 through March 2022. These are the dates for which forecasts were made and evaluated. A horizontal dashed line at 15,000 shows the hypothetical resource constraint $K$. Panel B shows allocation scores (AS) for each model's 14 day-ahead forecast, across all US states. The $x$-axis corresponds to the date of the observation that a model's prediction was targeting (e.g., the date the forecast was made plus the forecast horizon). AS typically are high when the observed value is near to the constraint, which occurs during the last Monday in December (on the way up) and the last Monday in January (on the way down). In Panel B, the \texttt{per-capita} allocation is drawn in a heavier solid line. Panel C shows the MWIS metric across weeks, averaged across all states. MWIS values tend to scale with the observed and predicted values, and the peak MWIS values happen around and just after the peak of the Omicron wave.}\label{fig:metrics-over-time}
\end{figure}

@@

Averaging across all weeks, the ensemble forecast achieved a better mean allocation score (MAS) than two benchmark
methods. The \texttt{COVIDhub-ensemble} had the best MAS across all weeks, the \texttt{per-capita} allocation had the second best MAS, and
the \texttt{COVIDhub-baseline} model had the sixth best allocation (Table \ref{tab:multi-week-performance-summary}). No
individual model had a better MAS than the per-capita allocation.

Model rankings on mean allocation score (MAS) and time-averaged mean weighted interval score
(taMWIS) were broadly similar, with some places of disagreement. The four most accurate and the five least accurate models were the same
according to both metrics, although not in exactly the same order (Table \ref{tab:multi-week-performance-summary}).
(This comparison excludes the \texttt{per-capita} allocation which does not use forecasts and therefore cannot be assigned a MWIS.)
Notably, the \texttt{JHUAPL-SLPHospEns} model ranked best on taMWIS but had only the fourth best MAS.

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Sat Feb 17 02:23:03 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
model & MAS & taMWIS & taMWIS rank \\ 
  \hline
COVIDhub-ensemble & 389 & 70 & 2 \\ 
  per-capita & 464 & - & - \\ 
  COVIDhub-trained\_ensemble & 483 & 87 & 4 \\ 
  CU-select & 502 & 81 & 3 \\ 
  JHUAPL-SLPHospEns & 526 & 67 & 1 \\ 
  COVIDhub-baseline & 594 & 93 & 5 \\ 
  JHUAPL-Bucky & 643 & 112 & 7 \\ 
  MUNI-ARIMA & 707 & 116 & 8 \\ 
  JHUAPL-Gecko & 929 & 110 & 6 \\ 
  USC-SI\_kJalpha & 1473 & 155 & 9 \\ 
   \hline
\end{tabular}
\caption{Time averaged mean weighted interval scores (taMWIS) and mean allocation scores (MAS) by model across 13 weeks. These results show the average performance across time for the nine models that submitted forecasts for every week from 2021-11-29 through 2022-02-21, as well
  as for a per-capita allocation. MWIS is not defined for the \texttt{per-capita} allocation which does not use forecasts.} 
\label{tab:multi-week-performance-summary}
\end{table}

@@

AS was computed for a range of $K$ from 200 to 60,000 for forecasts made on December 20, 2020 predicting levels of
hospitalizations on January 3, 2021 as well as for the \texttt{per-capita} allocation (Figure \ref{fig:multi-k}A). These
calculations highlight that AS was highest at at values of $K$ near the observed nationwide total hospital admissions of
19,581 that day. Model ranks by AS were fairly stable across all $K$, and there appears
to be an substantial interval around the observed value over which ranks are constant.

@@

Our example application of the AS refers to a hypothetical and unspecified resource need that is supposed to 
pair with our forecasted outcome (new hospital admissions) in a manner that might appear overly simplistic and 
artificially direct.
A concrete decision problem, which motivated our work on allocation scoring, is

allocation of ventilators during a respiratory viral pandemic
\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}.
